Spatio-angular fluorescence microscopy I. Basic theory (Draft: \today)

### 2018-11-07 papers/basic-theory/paper/paper.tex
Corrected a few typos.

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling an aplanatic microscope imaging an ensemble
  of in-focus fluorescent dipoles as a linear Hilbert-space operator with domain
  $\mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ and range $\mbb{L}_2(\mbb{R}^2)$, and we
  express the operator in terms of four different sets of object-space basis
  functions. We show that the operator takes a particularly convenient form when
  expressed in a basis of spatial and spherical harmonics---a form we call the
  spatio-angular dipole transfer function. We demonstrate our formalism by
  analyzing a single-view fluorescence microscope without using the monopole or
  scalar approximations. We show that this microscope has an angular band limit,
  and we demonstrate the value of the transfer function approach by efficiently
  simulating the imaging process with simple phantoms. Notably, we show that
  information about the out-of-plane orientation of ensembles of in-focus
  fluorophores is recorded by fluorescence microscopes, and we show that these
  effects persist in the paraxial limit. We discuss the implications of our
  analysis for all quantitative fluorescence microscopy studies and lay out a
  path towards a complete theory.
\end{abstract*}

@@

We start by defining \textit{pupil functions} for both monopole and dipole
imaging systems, and we explicitly relate the pupil function to the previously
defined coherent transfer function. Next, we calculate the transfer functions
for monopole and dipole imaging systems with high-NA objectives. We resort to
numerics as necessary, but we can calculate several transfer function in closed
form. We compare these results with paraxial results which are completely in
closed form.

@@

Since the second lens is paraxial, we can model with relationship between the
field in the aperture plane and the field on the detector with a scaled Fourier
transform \cite{goodman1996}
\begin{align}
   U_d(\rd', \ro) \propto \int_{\mbb{R}^2}d\rp\, p(\rp)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0} \rp\cdot\ro \right]\text{exp}\left[-2\pi i \frac{n_1}{\lambda f_1} \rp\cdot\rd' \right]. \label{eq:detectorfourier}
\end{align}
If we define $P(\bvp)$ as the two-dimensional Fourier transform of the pupil function
then we can rewrite Eq. \ref{eq:detectorfourier} as
\begin{align}
  U_d(\rd', \ro) \propto P\left(\frac{n_0}{\lambda f_0}\ro + \frac{n_1}{\lambda f_1}\rd'\right),
\end{align}
which we can simplify further in terms of $m = -\frac{f_1n_0}{f_0n_1}$
\begin{align}
  U_d(\rd' - m\ro) \propto P\left(\frac{n_1}{\lambda f_1}[\rd' - m\ro]\right),
\end{align}
The irradiance on the detector is given by the absolute square of the field so
\begin{align}
  h'(\rd' - m\ro) \propto \left|P\left(\frac{n_1}{\lambda f_1}[\rd' - m\ro]\right)\right|^2.
\end{align}
If we demagnify the coordinates $\rd = \rd'/m$ and the irradiance
$h(\rd - \ro) \propto h'(m[\rd - \ro])$ we find that the monopole point spread function
is related to the Fourier transform of the monopole pupil function by
\begin{align}
  h(\rd - \ro) \propto \left|P\left(-\frac{n_o}{\lambda f_o}[\rd - \ro]\right)\right|^2.
\end{align}
The monopole coherent spread function is the absolute square of the monopole point spread function so
\begin{align}
  c(\rd - \ro) \propto P\left(-\frac{n_o}{\lambda f_o}[\rd - \ro]\right).
\end{align}
The monopole coherent transfer function is the Fourier transform of the monopole coherent spread function so
\begin{align}
  C(\bvp) \propto p\left(\frac{\lambda f_o}{n_o}\bvp\right). \label{eq:monopupil}
\end{align}
Equation \ref{eq:monopupil} is the key result of this section---the monopole
coherent transfer function is a scaled monopole pupil function.

\subsection{Dipole pupil function} 
We define the \textit{dipole pupil function} $\mb{p}(\rp, \so)$ of the imaging
system as the electric field immediately following the aperture plane created by
a dipole oriented along $\so$. Since dipoles emit vector-valued electric fields,
the dipole pupil function is a vector-valued function. Almost all of the
arguments in the previous section carry over to the dipole case. Briefly, we can
write the electric field created at a point in the aperture $\rp$ by a dipole at
$\ro$ oriented along $\so$ as
 \begin{align}
   \mb{E}(\rp, \ro, \so) = \mb{p}(\rp, \so)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0} \rp\cdot\ro \right]. \label{eq:pupil2}
 \end{align}
 The second lens is paraxial, so we can find the field on the detector with a
 Fourier transform
 \begin{align}
   \mb{E}_d(\rd', \ro, \so) \propto \int_{\mbb{R}^2}d\rp\, \mb{p}(\rp, \so)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0} \rp\cdot\ro \right]\text{exp}\left[-2\pi i \frac{n_1}{\lambda f_1} \rp\cdot\rd' \right]. \label{eq:detectorfourier2}
\end{align}
Note that the Fourier transform of a vector field is the Fourier transform of
its scalar-valued orthogonal components, so Eq. \ref{eq:detectorfourier2}
specifies two two-dimensional Fourier transforms. We follow the same
manipulations as the previous section and find that the dipole coherent transfer
function is a scaled dipole pupil function
\begin{align}
  \mb{C}(\bvp, \so) \propto \mb{p}\left(\frac{\lambda f_o}{n_o}\bvp, \so\right). \label{eq:monopupil}
\end{align}

>>>>

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling an aplanatic microscope imaging an ensemble
  of in-focus fluorescent dipoles as a linear Hilbert-space operator with domain
  $\mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ and range $\mbb{L}_2(\mbb{R}^2)$, and we
  express the operator in terms of four different sets of object-space basis
  functions. We show that the operator takes a particularly convenient form when
  expressed in a basis of spatial and spherical harmonics---a form we call the
  spatio-angular dipole transfer function. We demonstrate our formalism by
  analyzing a single-view fluorescence microscope without using the monopole or
  scalar approximations. We show that this imaging system has an angular band
  limit, and we demonstrate the value of the transfer function approach by
  efficiently simulating the imaging process with simple phantoms. Notably, we
  show that information about the out-of-plane orientation of ensembles of
  in-focus fluorophores is recorded by fluorescence microscopes, and we show
  that these effects persist in the paraxial limit. We discuss the implications
  of our analysis for all quantitative fluorescence microscopy studies and lay
  out a path towards a complete theory.
\end{abstract*}

@@

We start by defining \textit{pupil functions} for both monopole and dipole
imaging systems, and we explicitly relate the pupil functions to the previously
defined coherent transfer functions. Next, we calculate the transfer functions
for monopole and dipole imaging systems with high-NA objectives. We resort to
numerics as necessary, but we can calculate several transfer function in closed
form. We compare these results with paraxial results which are completely in
closed form.

@@

Since the second lens is paraxial, we can model with relationship between the
field in the aperture plane and the field on the detector with a scaled Fourier
transform \cite{goodman1996}
\begin{align}
   U_d(\rd', \ro) \propto \int_{\mbb{R}^2}d\rp\, p(\rp)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0} \rp\cdot\ro \right]\text{exp}\left[-2\pi i \frac{n_1}{\lambda f_1} \rp\cdot\rd' \right]. \label{eq:detectorfourier}
\end{align}
If we define $P(\bvp)$ as the two-dimensional Fourier transform of the pupil function
then we can rewrite Eq. \ref{eq:detectorfourier} as
\begin{align}
  U_d(\rd', \ro) \propto P\left(\frac{n_0}{\lambda f_0}\ro + \frac{n_1}{\lambda f_1}\rd'\right),
\end{align}
which we can simplify further in terms of $m = -\frac{f_1n_0}{f_0n_1}$
\begin{align}
  U_d(\rd' - m\ro) \propto P\left(\frac{n_1}{\lambda f_1}[\rd' - m\ro]\right),
\end{align}
The irradiance on the detector is given by the absolute square of the field so
\begin{align}
  h'(\rd' - m\ro) \propto \left|P\left(\frac{n_1}{\lambda f_1}[\rd' - m\ro]\right)\right|^2.
\end{align}
If we demagnify the coordinates $\rd = \rd'/m$ and the irradiance
$h(\rd - \ro) \propto h'(m[\rd - \ro])$ we find that the monopole point spread function
is related to the Fourier transform of the monopole pupil function by
\begin{align}
  h(\rd - \ro) \propto \left|P\left(-\frac{n_o}{\lambda f_o}[\rd - \ro]\right)\right|^2.
\end{align}
The monopole coherent spread function is the absolute square of the monopole point spread function so
\begin{align}
  c(\rd - \ro) \propto P\left(-\frac{n_o}{\lambda f_o}[\rd - \ro]\right).
\end{align}
The monopole coherent transfer function is the Fourier transform of the monopole coherent spread function so
\begin{align}
  C(\bvp) \propto p\left(\frac{\lambda f_o}{n_o}\bvp\right). \label{eq:monopupil0}
\end{align}
Equation \ref{eq:monopupil0} is the key result of this section---the monopole
coherent transfer function is a scaled monopole pupil function.

\subsection{Dipole pupil function} 
We define the \textit{dipole pupil function} $\mb{p}(\rp, \so)$ of the imaging
system as the electric field immediately following the aperture plane created by
an on-axis dipole oriented along $\so$. Since dipoles emit vector-valued
electric fields, the dipole pupil function is a vector-valued function. Almost
all of the arguments in the previous section carry over to the dipole case.
Briefly, we can write the electric field created at a point in the aperture
$\rp$ by a dipole at $\ro$ oriented along $\so$ as
 \begin{align}
   \mb{E}(\rp, \ro, \so) = \mb{p}(\rp, \so)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0} \rp\cdot\ro \right]. \label{eq:pupil2}
 \end{align}
 The second lens is paraxial, so we can find the field on the detector with a
 Fourier transform
 \begin{align}
   \mb{E}_d(\rd', \ro, \so) \propto \int_{\mbb{R}^2}d\rp\, \mb{p}(\rp, \so)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0} \rp\cdot\ro \right]\text{exp}\left[-2\pi i \frac{n_1}{\lambda f_1} \rp\cdot\rd' \right]. \label{eq:detectorfourier2}
\end{align}
Note that the Fourier transform of a vector field is the Fourier transform of
its scalar-valued orthogonal components, so Eq. \ref{eq:detectorfourier2}
specifies two two-dimensional Fourier transforms. We follow the same
manipulations as the previous section and find that the dipole coherent transfer
function is a scaled dipole pupil function
\begin{align}
  \mb{C}(\bvp, \so) \propto \mb{p}\left(\frac{\lambda f_o}{n_o}\bvp, \so\right). \label{eq:monopupil}
\end{align}
